---
title: "DTSA_5301_COVID_Final_Project"
author: "Bryan P. Mayfield"
date: '2022-06-15'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading packages, include=FALSE}

library(tidyverse)
library(lubridate)

```

```{r, reading in data, include=FALSE}

confirmed_us <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")

confirmed_global <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")

deaths_us <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")

deaths_global <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")

lookup_table <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv")
```

# Questions of Interest for this Analysis 

The question of interest in this analysis is twofold, have some areas (states) handled the pandemic better than others as it relates to the number of cases and deaths? And two, what factors contribute to the number of deaths (measure as deaths per thousand)?

# Data Source and Description 

This data is from the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19). The upstream data sources are listed in the README of the Johns Hopkins repository. The data used in this analysis contains both case and deaths information for both the US individually as well as global data.

# Data Exploration and Cleaning

```{r}

head(confirmed_us)
```

```{r}

head(confirmed_global)
```

```{r}

head(confirmed_us)
```

```{r}

head(deaths_global)
```

```{r}

head(deaths_us)
```

```{r}

confirmed_global_clean <- confirmed_global %>% 
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
               names_to = "date",
               values_to = "cases") %>% 
  select(-c(Lat, Long))

head(confirmed_global_clean)
```

```{r}

deaths_global_clean <- deaths_global %>% 
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
               names_to = "date",
               values_to = "deaths") %>% 
  select(-c(Lat, Long))


head(deaths_global_clean)
```

```{r}

global_combined <- confirmed_global_clean %>% 
  full_join(deaths_global_clean) %>% 
  rename(country_region = `Country/Region`,
         province_state = `Province/State`) %>% 
  mutate(date = mdy(date))

head(global_combined)
```

```{r}

summary(global_combined)

```

```{r}

confirmed_us_clean <- confirmed_us %>% 
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>% 
  select(Admin2:cases) %>% 
  mutate(date = mdy(date)) %>% 
  select(-c(Lat, Long_)) %>% 
  rename(province_state = Province_State,
         country_region = Country_Region)

head(confirmed_us_clean)
```

```{r}

deaths_us_clean <- deaths_us %>% 
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>% 
  select(Admin2:deaths) %>% 
  mutate(date = mdy(date)) %>% 
  select(-c(Lat, Long_)) %>% 
  rename(province_state = Province_State,
         country_region = Country_Region)

head(deaths_us_clean)
```

```{r}

us_combined <- confirmed_us_clean %>% 
  full_join(deaths_us_clean)

head(us_combined)
```

```{r}

global_combined_2 <- global_combined %>% 
  unite("Combined_Key",
        c(province_state, country_region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE) 

head(global_combined_2)
```

```{r}

lookup_table_clean <- lookup_table %>% 
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2)) %>% 
  rename(province_state = Province_State,
         country_region = Country_Region)

head(lookup_table_clean)

```

```{r}

global_combined_3 <- global_combined_2 %>% 
  left_join(lookup_table_clean, by = c("province_state", "country_region")) %>% 
  select(-c(UID, FIPS)) %>% 
  select(province_state, country_region, date, cases, deaths, Population, Combined_Key)

head(global_combined_3)
```

# Analysis and Visualization 

```{r}

us_by_state <- us_combined %>% 
  group_by(province_state, country_region, date) %>% 
  summarize(cases = sum(cases),
            deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_million = deaths * 1000000 / Population) %>%
  select(province_state, country_region, date, cases, deaths, deaths_per_million, Population) %>%
  ungroup()

head(us_by_state)
```

```{r}


us_totals <-  us_by_state %>% 
  group_by(country_region, date) %>% 
  summarise(cases = sum(cases),
            deaths = sum(deaths),
            Population = sum(Population)) %>% 
  mutate(deaths_per_million = deaths * 1000000 / Population) %>% 
  select(country_region, date, cases, deaths, deaths_per_million, Population) %>% ungroup()

head(us_totals)
```

```{r}

tail(us_totals)
```

```{r}

us_totals %>% 
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in the United States", y = NULL)



```

```{r}

state_1 = "Texas"

state_2 = "Florida"

us_by_state %>% 
  filter(province_state == state_1) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state_1), y = NULL)


```

```{r}

us_by_state %>% 
  filter(province_state == state_2) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state_2), y = NULL)

```

```{r}

max(us_totals$deaths)
```

```{r}

us_by_state_2 <- us_by_state %>% 
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))


```

```{r}

us_totals_2 <-  us_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

tail(us_totals_2 %>% select(new_cases, new_deaths, everything()))
```

It is surprising that the number of new deaths on June 12th, 2022 is negative. This may be some sort of error in the data collection?

```{r}

us_totals_2 %>% 
  ggplot(aes(x= date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color = "new_cases")) + 
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID 19 in the United States", y = NULL)
```

This visualization (above) seems to provide a bit more insight into the trajectory of cases. The spike in early 2022 may be related to the Omicron variant.

```{r}

us_by_state_2 %>% 
  filter(province_state == state_1) %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color = "new_cases")) + 
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state_1), y = NULL)

```

```{r}

us_by_state_2 %>% 
  filter(province_state == state_2) %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color = "new_cases")) + 
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state_2), y = NULL)

```

```{r}

us_state_totals <- us_by_state_2 %>% 
  group_by(province_state) %>% 
  summarise(deaths = max(deaths), 
            cases = max(cases),
            population_ = max(Population),
            cases_per_thousand = 1000 * cases / population_,
            deaths_per_thousand = 1000 * deaths / population_)

head(us_state_totals)
```

```{r}

us_state_totals %>% 
  slice_min(deaths_per_thousand, n = 10)
```

Although the above table is interesting, it doesn't address the fact that many places with a small number of deaths and/or cases are also places with much smaller populations. A better measure of how well places are doing would be to measure cases and deaths as a proportion of the overall population of that area (`province_state`) as opposed to cases or deaths per thousand.

```{r}

us_state_totals %>% 
  slice_max(deaths_per_thousand, n = 10)

```

While it is interesting that Grand Princess has a population of 0, it should be filtered out.

```{r}

us_state_totals %>% 
  slice_max(deaths_per_thousand, n = 10) %>% 
  filter(population_ > 0)
```

# Modeling

In the visualizations, I intentionally decided to keep 0 values, but this will lead to a problem, so here I will create a data set that filters those observations out for modeling purposes. Given more time, it might be prudent to further investigate observations with 0 values, especially those with 0 values for population (like Grand Princess, which showed a population of 0, but had 3 deaths).

```{r}

us_state_totals_modeling <- us_state_totals %>% 
  filter(population_ > 0)
```

```{r}

lecture_model <- lm(deaths_per_thousand ~ cases_per_thousand, data = us_state_totals_modeling)

summary(lecture_model)
```

```{r}

us_state_totals_modeling_w_preds <- us_state_totals_modeling %>% 
  mutate(pred = predict(lecture_model))

head(us_state_totals_modeling_w_preds)
```

```{r}

us_state_totals_modeling_w_preds %>% 
  ggplot() +
  geom_point(aes(x = cases_per_thousand, y = deaths_per_thousand), color = "blue") +
  geom_point(aes(x = cases_per_thousand, y = pred), color = "red")

```

Let's see if we can improve this model by adding variables/features.

```{r}

second_model <- lm(deaths_per_thousand ~ population_ + cases_per_thousand, data = us_state_totals_modeling)

summary(second_model)
```

```{r}

us_state_totals_modeling_w_second_preds <- us_state_totals_modeling %>% 
  mutate(pred_2 = predict(second_model))

head(us_state_totals_modeling_w_second_preds)
```

```{r}

us_state_totals_modeling_w_second_preds %>% 
  ggplot() +
  geom_point(aes(x = cases_per_thousand, y = deaths_per_thousand), color = "blue") +
  geom_point(aes(x = cases_per_thousand, y = pred_2), color = "red")


```

We don't really see much of an improvement here as expected from the model summary showing that `population_` is not a statistically significant contributor to the model. Given more time, we could explore the data set for more features that might explain the changes in \`deaths_per_thousand.

# Conclusion and Possible Sources of Bias 

In conclusion, in this analysis we were able to explore some of the differences in the number of COVID19 cases and death by both country and state. We took a deeper dive into differences between states and for my particular analysis I looked at two states in which I lived during the pandemic, Texas and Florida. We could see that some states handled the pandemic better than others in terms of number of cases and deaths normalized by population.

However, it is important to note that there are possible sources of bias. For example, one possible source that has received media attention is the way in which deaths are reported. Some coroners in some areas may report and attribute a COVID19 death to something other than COVID19 such as pneumonia or upper respiratory illness or multi organ failure in an effort to decrease the visibility into the actual impact of COVID19 on that particular area.

Another possible source of bias is in the number of cases variable. There are potentially many undiagnosed or asymptomatic cases that likely were not reported in any given area. It is unclear what the prevalence of this might have been so it would be quite difficult to eliminate this source of bias.
