---
title: "NYPD Shooting Incident Data (Historic)"
author: "MSDS_student"
date: '2022-05-03'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-   Check for missing data

```{r loading_packages}

library(tidyverse)
library(lubridate)
library(janitor)

```

```{r importing_data}

nypd_shooting_incident_data <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")


```

```{r viewing_data}

nypd_shooting_incident_data


```

This is historical data containing information regarding individual shooting incidents that occurred from 2006 to the end of the previous calendar year.

```{r summary_of_data}

summary(nypd_shooting_incident_data)

```

```{r tidying_cleaning}


nypd_shooting_incident_data_clean <-nypd_shooting_incident_data %>%
  clean_names() 

nypd_shooting_incident_data_clean <- nypd_shooting_incident_data_clean %>%
  mutate(occur_date = mdy(occur_date)) %>%
  mutate(boro = as_factor(boro)) %>%
  mutate(perp_sex = as_factor(perp_sex)) %>%
  mutate(perp_age_group = as_factor(perp_age_group)) %>%
  mutate(perp_age_group = fct_relevel(perp_age_group, "<18", "18-24", "25-44", "45-64", "65+", "UNKNOWN")) %>%
  mutate(vic_age_group = as_factor(vic_age_group)) %>%
  mutate(vic_sex = as_factor(vic_sex)) %>%
  mutate(perp_race = as_factor(perp_race)) %>%
  mutate(vic_race = as_factor(vic_race)) %>%
  select(-c(incident_key, jurisdiction_code, location_desc, x_coord_cd, y_coord_cd, latitude, longitude, lon_lat))
  




```

Above I have converted the class of multiple variables to date objects or factors depending on the variable. Although, the location data could be interesting and result in some interesting insights and data visualizations, it is out of the scope of this particular project/analysis so I have dropped these columns from the data set.

Although the summary function provided some useful information, it still is not super clear where we have missing values. Next I will check relevant columns for missing values (for this analysis, I am not particularly concerned with missing values for `occur date` or `occur_time)`.

```{r identify_missing_values}


# Use sum(is.na) to check for and count the number of missing values in each column

is.na(nypd_shooting_incident_data_clean) %>%
  colSums() 


```

From this analysis step to identify missing values, it looks like we are missing values in 3 columns, `perp_age_group`, `perp_sex`, and `perp_race`. Given unlimited time, I could likely identify some more elegant ways to deal with these missing values, but for the purpose of this simple analysis, it will be sufficient to just drop them.

```{r remove_NA_values}

nypd_shooting_incident_data_clean <- nypd_shooting_incident_data_clean %>%
  drop_na()

```

One interesting analysis question might be is there a difference in the number of incidents by neighborhood.

```{r difference_by_boro}

nypd_by_boro <- nypd_shooting_incident_data_clean %>%
  group_by(boro) %>%
  summarize(number_of_incidents = n())




```

This numerical data is interesting and it shows that certain neighborhoods do seem to have more incidents than others. It may be easier to see the relationship if it is a visualization.

```{r bar_plot_by_boro}


nypd_shooting_incident_data_clean %>%
  group_by(boro) %>%
  summarize(number_of_incidents = n()) %>%
  ggplot(aes(x = fct_reorder(boro, number_of_incidents), y = number_of_incidents)) + 
  geom_col()

```

The above plot makes the point a bit clearer regarding the difference in incidents by neighborhood. Below are some additional interesting visualizations.

```{r percentage_murders}

nypd_shooting_incident_data_clean %>%
  ggplot(aes(x = statistical_murder_flag)) +
  geom_bar()

```

```{r bar_plot_race}

# Bar chart looking at the race of the perpetrators 

nypd_shooting_incident_data_clean %>%
  mutate(perp_race = perp_race %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(x = perp_race)) + 
  geom_bar() + 
  coord_flip()


```

```{r bar_plot_age_group}


nypd_shooting_incident_data_clean %>%
  ggplot(aes(x = perp_age_group)) + 
  geom_bar()


```

```{r bar_plot_sex_race}

nypd_shooting_incident_data_clean %>%
  ggplot(aes(x = vic_sex, fill = vic_race)) + 
  geom_bar()
```

```{r model_incidents_boro}

lm(number_of_incidents ~ boro + 0, data = nypd_by_boro)



```

```{r model_multivariate}

lm(boro ~ perp_race + vic_race, data = nypd_shooting_incident_data_clean)


```

## Conclusion 

This brief EDA showed some interesting differences in the number of incident occurrences among different neighborhoods, age groups, and races. Further exploration of this data set and more sophisticated analysis would likely lead to some explanations for these differences. In terms of bias, it is possible that there may be some reporting bias since we aren't completely sure how this data was collected as well as the goals or objectives of those performing or directing the data collection. Additionally, the large amount of NA rows that had to be removed might have shifted the results some. In terms of personal bias, my own age group, race, and experiences with neighborhoods in New York City may have impacted the questions I asked of the data. In order to mitigate that, I tried to make the most objective observations possible without weighting the data in any particular direction and I remained aware of my own biases throughout the course of the analysis so that I could constantly question if that bias might be having any impact on the objectiveness of the analysis.
